---
title: "Creating Graphics with R"
author: "Patrick Chester at NYU Data Services: data.services@nyu.edu"
date: '`r format(Sys.time(), "%B %d %Y")`'
output: html_document
---

Last edited: 8/22/2019


# I. Base Graphics -

```{r}
library(MASS)
data(UScereal)
head(UScereal)
```



## A. Choosing the Right Graph for your Data

### One Continuous Variable
Key questions:
- What is the Distribution of Variable X?
- Is my Variable X normally distributed? Bimodal? Skewed?
- Are there any outliers in my variable?

Histogram

```{r}
hist(UScereal$calories, breaks = 15)
```

Boxplot

```{r}
boxplot(UScereal$calories, horizontal = TRUE)
```



### One Categorical (Discrete) Variable

Key question(s):
- Is the distribution across categories of Variable X evenly distributed?

```{r}
barplot(table(UScereal$shelf))
```




### Two Continuous Variables

Key questions:
- Is there a relationship between Variable X and Variable Y?
- If there is a relationship, is it linear? Quadratic? 

Scatterplot
```{r}
plot(x = UScereal$sugars, y = UScereal$calories)
```

Using a formula to create a scatterplot
```{r}
plot(calories ~ sugars, data = UScereal) # formula notation
```

Scatterplot matrix
```{r}
plot(UScereal[, c(2:8, 10)]) # scatterplot matrix
```




### One Continuous Variable and One Categorical Variable

Key question(s):
- Is the distribution of Variable Y, different across categories of Variable X?

```{r}
boxplot(sugars ~ shelf, data = UScereal)
```




### Two Continuous Variables and One Categorical Variable

Key questions:
- Is the relationship between Variable X and Y different across categories of Variable Z?

```{r}
plot(calories ~ sugars, data = UScereal, col = shelf)
```



## B. Graphing Elements

### Titles

*Tip: We want titles to meaningful, non-repetitive and include units when applicable*

Use the ylab, xlab and main arguments within the plot command

```{r}
plot(calories ~ sugars, data = UScereal, ylab = 'Calories',
     xlab = 'Sugars (grams)', main = 'Nutrition of a Single Cup of Cereal')
```


Use the title function after the plot has been generated
If using the title function, the ann argument inside plot must be set to FALSE

```{r}
plot(calories ~ sugars, data = UScereal, ann = FALSE)
title(main = 'Nutrition of a Single Cup of Cereal', ylab = 'Calories',
      xlab = 'Sugars (grams)') # add afterwards
```




### Legend

Use the legend function after the plot has been generated

```{r}
plot(calories ~ sugars, data = UScereal, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('green', 'red', 'black'), title = "Shelf")
```



### Point Shape and Color 

Tip: Changing color or shape of points can be used to represent the same dimension

The pch argument changes the shape of the points

```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
```



Set a color to a factor variable, and R will use default colors
```{r}
plot(calories ~ sugars, data = UScereal, pch = 19, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('green', 'red', 'black'), title = "Shelf")
```


Use a palette of defined colors
```{r}
palette(c('#e5f5f9', '#99d8c9', '#2ca25f'))
plot(calories ~ sugars, data = UScereal, pch = 19, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('#e5f5f9', '#99d8c9', '#2ca25f'), title = "Shelf")
```


Label points with the text function

```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
text(UScereal$sugars, UScereal$calories, row.names(UScereal),
     col = "red", pos = 1, cex = .5)
```


The pos argument in the text funciton changes the position of the labels
```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
text(UScereal$sugars, UScereal$calories, UScereal$mfr, 
     col = "blue", pos = 2)
```


### Identify Outliers

1. Create a subset of datas which are the outliers, and then use that subset to label

```{r}
plot(calories ~ sugars, data = UScereal, pch = 19)
outliers <- UScereal[which(UScereal$calories > 300),]
text(outliers$sugars, outliers$calories, labels = row.names(outliers),pos = 1)
```

2. Or remove them by subsetting UScereal
```{r}
plot(calories ~ sugars, data = UScereal[UScereal$calories < 300, ], pch = 19)
```

### Size (text size, point size, label size etc..)

Use the cex argument family to change the size of any attribute
```{r}
plot(calories ~ sugars, data = UScereal, pch = 19, ann = FALSE, cex = 1.5)
outliers <- UScereal[which(UScereal$calories > 300),]
text(outliers$sugars, outliers$calories, pos = 1,
     labels = row.names(outliers), cex = .75)
title(main = 'Nutrition of a Single Cup of Cereal', ylab = 'Calories',
      xlab = 'Sugars (grams)', cex.main = 2, cex.lab = 1.5)
```


Combine Graphs into the Same Window
```{r}
par(mfrow = c(2, 2))

boxplot(calories ~ shelf, data = UScereal)
hist(UScereal$calories, breaks = 15)
boxplot(sugars ~ shelf, data = UScereal)
hist(UScereal$sugars, breaks = 15)

par(mfrow = c(1, 1)) # reset the matrix
```


### Exercise 1a. 
How can we improve this graph? From what we have learned above, implement at least 3 improvements to this graph.
A little more information about this dataset: The heart and body weights of samples of male and female cats used for digitalis experiments. The cats were all adult, over 2 kg body weight.

```{r}
data(cats)
head(cats)
plot(Hwt~Bwt, data = cats)
```

### Exercise 1b.
How can we improve this graph? From what we have learned above, implement at least 3 improvements to this graph. A little more information about this dataset: The heart and body weights of samples of male and Fisher's famours dataset measures the sepal and petal length and width for 3 species of Iris
```{r}
data(iris)
head(iris)
boxplot(Petal.Width ~ Species, data = iris)
```


# II. Advanced Graphics using ggplot2

```{r}
#install.packages('ggplot2')
library(ggplot2)
```

## A. The ggplot function

Base q plot function
```{r}
qplot(x = sugars, y = calories, color = as.factor(shelf), data = UScereal) 
```

Use the ggplot function to get started
```{r}
p1 <- ggplot(UScereal, aes(x = calories))
```


## B. Layers 

Add layers to the original ggplot object with a '+'
```{r}
p1 + geom_dotplot()

p1 + geom_density()

p1 + geom_histogram(binwidth = 10)
```

And we add more layers
```{r}
p1 + geom_histogram(binwidth = 10) + 
  xlab('Calories in a Single Cup') +
  ylab('Frequency') + 
  ggtitle('Distribution of Calories in US Cereals') + 
  theme_minimal()
#theme_bw()
```

The order of the layers does not matter
```{r}
p1 + geom_histogram(binwidth = 10) + 
  xlab('Calories in a Single Cup') +
  ylab('Frequency') + 
  ggtitle('Distribution of Calories in US Cereals') + 
  theme_bw() + 
  theme(text = element_text(size = 20))
```

You can also add multiple geom_function layers to the same graph
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories, color = mfr))
```

Overlaying lines that connect poitns
```{r}
p2  + geom_point() + 
  geom_line()
```

```{r}
p2  + geom_point() + 
  geom_smooth(color ="gray", method = "lm", se = FALSE)

p2  + geom_point() + 
  geom_smooth(color ="gray", method = "loess", se = TRUE)
```

## C. Aesthetics: x position, y position, size of elements, shape of elements, color of elements 
elements: geometric shapes such as points, lines, line segments, bars and text
geomitries have their own aesthetics i.e. points have their own shape and size


To color by manufacturer - we can put the color in the ggplot function:
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories, color = mfr))

p2 + geom_point() 
```

Or inside the geom_point function:
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories))

p2 + geom_point(aes(color = mfr)) 
```

We can also scale colors manually using the appropriately named scale_color_manual function
```{r}
my_colors <- c('#9ebcda', '#8c96c6', '#8c6bb1', '#88419d', '#810f7c', '#4d004b')

p2 + geom_point() + 
  scale_color_manual(values = my_colors) 
```

### Adding Labels to points

Use the geom_text() layer
```{r}
p2 + geom_point(aes(color = mfr)) + 
  geom_text(aes(label = row.names(UScereal)), hjust = 1.1)
```

Changing point size
```{r}
p2 + geom_point(aes(color = mfr), size = 4) 
```


### Editing the legend

Use the scale_color_manual() layer, and the color argument in the labs() layer 
```{r}
p2 + geom_point(aes(color = mfr), size = 2) + 
  labs(color = 'Manufacturer') + 
  scale_color_manual(values = c('blue', 'green', 'purple', 'navyblue', 'red', 'orange'),  
                     labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina')) + 
  theme(text = element_text(size = 15)) 
```

## D. Faceting  - divide a plot into subplots based on the valuesof one or more discrete variables

Tip: Use facets to help tell your story
Key questions:
- How is the distribution of sugar across different shelves?
- Are cereals with higher sugar content on lower shelves/at a child's eye level?

```{r}
p3 <- ggplot(UScereal, aes(x = sugars))

p3 + geom_histogram(binwidth = 4)
```

Each graph is in a separate row of the window
```{r}
p3 + geom_histogram(binwidth = 4) + 
  facet_grid(shelf ~ .)
```

Each graph is in a separate column of the window
```{r}
p3 + geom_histogram(binwidth = 4) + 
  facet_grid(. ~ shelf)
```

Finished product 
```{r}
p3 + geom_histogram(fill = '#3182bd', color = '#08519c', binwidth = 4) +
  facet_grid(shelf ~ .) + 
  theme(text = element_text(size = 15)) + 
  labs(title = 'Are Sugary Cereals on Lower Shelves?',
       x = 'Sugars (grams)', y = 'Count')

#html color. colormixer. Rcolorbrewer
```

### Box Plots
```{r}
p4 <- ggplot(UScereal, aes(mfr, calories))
```

```{r}
p4 + geom_boxplot()
```

```{r}
p4 + geom_boxplot(notch = TRUE)
```

```{r}
p4 + geom_violin()
```

```{r}
p4 + geom_boxplot(outlier.shape = 8, outlier.size = 3, fill = '#3182bd') + 
  coord_flip() + 
  labs(x = 'Manufacturer', y = 'Calories') + 
  theme_bw() + 
  scale_x_discrete(labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina'))
```

# Add median value to boxplot

```{r}
#install.packages('dplyr')
library(dplyr)
p4_meds <- UScereal %>% 
  group_by(mfr) %>% 
  summarise(med = round(median(calories), 1))
```

```{r}
p4 + geom_boxplot(outlier.shape = 8, outlier.size = 3, fill = '#8c96c6') + 
  labs(x = 'Manufacturer', y = 'Calories') + 
  theme_bw() + 
  scale_x_discrete(labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina')) + 
  geom_text(data = p4_meds, aes(x = mfr, y = med, label = med), size = 4, vjust = 1.4)
```


### Exercise 2a. Use ggplot2 to improve the graph below:
```{r}
data(cats)
head(cats)
plot(Hwt~Bwt, data = cats)
```
```{r}

p9<-ggplot(data=cats,aes(x=Bwt,y=Hwt,color=Sex))
p9 + geom_point(aes(color = Sex)) 
p8<-ggplot(data=cats,aes(x=Bwt))
p8 + geom_dotplot() +theme_classic()


```


### Exercise 2b. Use ggplot2 to improve the graph below:
```{r}
data(iris)
head(iris)
boxplot(Petal.Width ~ Species, data = iris)
```




# III. Other Graphics

## Scatterplot Matrix
```{r}
install.packages('GGally')
library(GGally)
ggpairs(UScereal[, c(2, 8, 9, 11)],
        upper = list(continuous = 'smooth', combo = 'facetdensity', discrete = 'blank') ,
        lower = list(continuous = 'cor', combo = 'box'))
```

Maps
http://bcb.dfci.harvard.edu/~aedin/courses/R/CDC/maps.html
http://rstudio.github.io/leaflet/
maps, choroplethr, 


# IV. Exercise Solutions

Below are just examples of cleaned up graphics. There are many solutions on how to improve these graphs

## Exercise 1a 

```{r}
palette(c('#fa9fb5', '#9ebcda'))
plot(Hwt~Bwt, data = cats, ylab  = 'Heart Weight (grams)', xlab = 'Body Weight (kg)', main = 'Measurements of Cats', pch = 16, col = Sex)
```

## Exercise 1b
```{r}
boxplot(Petal.Width ~ Species, data = iris, xlab = 'Petal Width (centimeters)', 
        main = 'Distribution of Petal Length by Species', pch = 8, horizontal = TRUE, col = 'lightgray')
```



## Exercise 2a
```{r}
e2a <- ggplot(cats, aes(Bwt, Hwt, color = Sex))
e2a + geom_point() + 
  theme_bw() + 
  labs(title = 'Measurements of Cats', x = 'Body Weight (kg)', y = 'Heart Weight (grams)')
```

## Exercise 2b
```{r}
e2b <- ggplot(iris, aes(Species, Petal.Width))
e2b + geom_boxplot(fill = '#addd8e') + 
  theme_bw() + 
  coord_flip() + 
  ggtitle('Distribution of Petal Length by Species' )
```


